---
title: 'Contact matrices'
teaching: 10
exercises: 2
---

:::::::::::::::::::::::::::::::::::::: questions 

- What is a contact matrix?
- How are contact matrices estimated?
- How are contract matrices used?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Objective 1
- Objective 2

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

## Prerequisites



:::::::::::::::::::::::::::::::::


## Introduction

Introductory text

:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor



::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::

## The contact matrix

Contact matrices can be estimated from surveys or contact data, or synthetic ones can be used.



## Estimating contact matrices

## Using contact matrices

### Measles example

### In mathematical models

Consider the SIR model where individuals are categorized as either susceptible $S$, infected but not yet infectious $E$, infectious $I$ or recovered $R$. The schematic below shows the processes which describe the flow of individuals between the disease states $S$, $I$ and $R$ and the key parameters for each process.

```{r diagram, echo = FALSE, message = FALSE}
DiagrammeR::grViz("digraph {

  # graph statement
  #################
  graph [layout = dot,
         rankdir = LR,
         overlap = true,
         fontsize = 10]

  # nodes
  #######
  node [shape = square,
       fixedsize = true
       width = 1.3]

       S
       I
       R

  # edges
  #######
  S -> I [label = ' infection \n(transmission rate &beta;)']
  I -> R [label = ' recovery \n(recovery rate &gamma;)']

}")
```

The [differential equations](../learners/reference.md#ordinary) below describe how individuals move from one state to another [(Bjørnstad et al. 2020)](https://doi.org/10.1038/s41592-020-0822-z).


$$
\begin{aligned}
\frac{dS}{dt} & = - \beta S I \\
\frac{dI}{dt} &= \beta S I - \gamma I \\
\frac{dR}{dt} &=\gamma I \\
\end{aligned}
$$
To add age structure to our model, we need to add additional equations for the infection states $S$, $I$ and $R$ for each age group $i$. If we want to assume that there is heterogeneity in contacts between age groups then we must adapt the transmission term $\beta SI$ to include the contact matrix $C$ as follows :

$$ \beta S_i \sum_j C_{i,j} I_j. $$ 
Susceptible individuals in age group $i$ become infected dependent on their rate of contact with individuals in each age group. For each disease state ($S$, $E$, $I$ and $R$) and age group ($i$), we have a differential equation describing the rate of change with respect to time.  

$$
\begin{aligned}
\frac{dS_i}{dt} & = - \beta S_i \sum_j C_{i,j} I_j \\
\frac{dI_i}{dt} &= \beta S_i\sum_j C_{i,j} I_j - \gamma I_i \\
\frac{dR_i}{dt} &=\gamma I_i \\
\end{aligned}
$$

::::::::::::::::::::::::::::::::::::: callout
### Check the dimension of $\beta$

In the SIR model without age structure the rate of contact is part of the transmission rate $\beta$, where as in the age-structured model we have separated out the rate of contact, hence the transmission rate $\beta$ in the age structured model will have a different value.

::::::::::::::::::::::::::::::::::::::::::::::::


### Contact groups

In the example above the dimension of the contact matrix will be the same as the number of age groups i.e. if there are 3 age groups then the contact matrix will have 3 rows and 3 columns. Contact matrices can be used for other groups as long as the dimension of the matrix matches the number of groups.

For example, we might have a meta population model with two geographic areas. Then our contact matrix would be a 2 x 2 matrix with entries representing the contact between and within the geographic areas.


## Normalisation

::::::::::::::::::::::::::::::::::::: callout
## Normalising matrices
Add additional maths (or epi) content for novice learners

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: callout
### Why would a contact matrix be non-symmetric?

One of the arguments of the function `contact_matrix()` is `symmetric=TRUE`. This means that the total number of contacts of age group 1 with age group 2, should be the same as the total number of contacts of age group 2 and age group 1 (see the `socialmixr` [vignette](https://cran.r-project.org/web/packages/socialmixr/vignettes/socialmixr.html) for more detail). However, when contact matrices are estimated from surveys or other sources, the *reported* number of contacts may differ by age group resulting in a non-symmetric contact matrix [(Prem et al 2021)](https://doi.org/10.1371/journal.pcbi.1009098).
::::::::::::::::::::::::::::::::::::::::::::::::







::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 1 : can the learner run existing code

Load contact and population data from socialmixr::polymod

```r
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 40),
  symmetric = TRUE
)
contact_data
```

:::::::::::::::::::::::: solution 

## Output
 
```{r polymod_uk, echo = FALSE}
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 40),
  symmetric = TRUE
)
contact_data
```


:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::


## Section header

Lesson content


::::::::::::::::::::::::::::::::::::: challenge 

## Challenge 2 : edit code/answer a question

Load contact and population data for Poland from socialmixr::polymod using the following age bins:

+ [0,15)
+ [15, 50)
+ 50 +

:::::::::::::::::::::::: solution 

```{r polymod_poland}
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "Poland",
  age.limits = c(0, 15, 50),
  symmetric = TRUE
)
contact_data
```
:::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::





::::::::::::::::::::::::::::::::::::: keypoints 

- Summarise the key points of the lesson using bullet points


::::::::::::::::::::::::::::::::::::::::::::::::
