---
title: 'Modelling disease burden'
teaching: 10 # teaching time in minutes
exercises: 5 # exercise time in minutes
---

:::::::::::::::::::::::::::::::::::::: questions 

- How can we model disease burden?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Understand when mathematical models of transmission can be separated from models of burden
- Generate disease burden using a burden model

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: prereq

+ Complete tutorial on [Simulating transmission](../episodes/simulating-transmission.md)

:::::::::::::::::::::::::::::::::


## Introduction

Introductory text


```{r, warning = FALSE, message = FALSE}
library(epiparameter)
library(epidemics)
library(socialmixr)
library(tidyverse)
```


:::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::: instructor

Inline instructor notes can help inform instructors of timing challenges
associated with the lessons. They appear in the "Instructor View".


::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::::


## Models of transmission and models of burden

Lesson content




## A burden model

```{r, echo = FALSE, message = FALSE}
# load contact and population data from socialmixr::polymod
polymod <- socialmixr::polymod
contact_data <- socialmixr::contact_matrix(
  polymod,
  countries = "United Kingdom",
  age.limits = c(0, 20, 40),
  symmetric = TRUE
)

# prepare contact matrix
contact_matrix <- t(contact_data$matrix)

# prepare the demography vector
demography_vector <- contact_data$demography$population
names(demography_vector) <- rownames(contact_matrix)

# initial conditions: one in every 1 million is infected
initial_i <- 1e-6
initial_conditions_inf <- c(
  S = 1 - initial_i, E = 0, I = initial_i, R = 0, V = 0
)

initial_conditions_free <- c(
  S = 1, E = 0, I = 0, R = 0, V = 0
)

# build for all age groups
initial_conditions <- rbind(
  initial_conditions_inf,
  initial_conditions_free,
  initial_conditions_free
)
rownames(initial_conditions) <- rownames(contact_matrix)

# prepare the population to model as affected by the epidemic
uk_population <- epidemics::population(
  name = "UK",
  contact_matrix = contact_matrix,
  demography_vector = demography_vector,
  initial_conditions = initial_conditions
)

# time periods
preinfectious_period <- 3.0
infectious_period <- 7.0
basic_reproduction <- 1.46

# rates
infectiousness_rate <- 1.0 / preinfectious_period
recovery_rate <- 1.0 / infectious_period
transmission_rate <- basic_reproduction / infectious_period

# run an epidemic model using `epidemic()`
output_plot <- epidemics::model_default(
  population = uk_population,
  transmission_rate = transmission_rate,
  infectiousness_rate = infectiousness_rate,
  recovery_rate = recovery_rate,
  time_end = 600, increment = 1.0
)

new_cases <- new_infections(output_plot,
                            by_group = FALSE)
```



```{r, message = FALSE}
# define delay parameters
onset_to_admission <- epiparameter(
  disease = "Disease X",
  epi_name = "onset to admission",
  prob_distribution = create_prob_distribution(
    prob_distribution = "gamma",
    prob_distribution_params = c(shape = 5, scale = 4),
    discretise = TRUE
  )
)


admission_to_discharge <- epiparameter(
  disease = "Disease X",
  epi_name = "admission to discharge",
  prob_distribution = create_prob_distribution(
    prob_distribution = "gamma",
    prob_distribution_params = c(shape = 10, scale = 2),
    discretise = TRUE
  )
)

x_range <- seq(0, 60, by = 0.1)
density_df <- data.frame(days = x_range,
                         density_admission = density(admission_to_discharge, x_range),
                         density_discharge = density(onset_to_admission, x_range))

ggplot(data = density_df, aes(x = days, y = density_admission))  +
  geom_line() +
  theme_bw() +
  labs(
    x = "onset to admission (days)",
    y = "pdf"
  )

ggplot(data = density_df, aes(x = days, y = density_discharge)) +
  geom_line() +
  theme_bw() +
  labs(
    x = "admission to discharge (days)",
    y = "pdf"
  )

```

```{r, message = FALSE}
ihr <- 0.1 # infection-hospitalisation ratio

# Calculate expected numbers with convolution:
hosp <- new_cases$new_infections * ihr

# define tail of the delay distribution
tail_value_admission <- quantile(onset_to_admission, 0.999)
tail_value_discharge <- quantile(admission_to_discharge, 0.999)

hospitalizations <- convolve(hosp, rev(density(onset_to_admission, 0:tail_value_admission)), type = "open")[1:length(hosp)]
discharges <- convolve(hospitalizations, rev(density(admission_to_discharge, 0:tail_value_discharge)), type = "open")[1:length(hospitalizations)]
in_hospital <- cumsum(hospitalizations) - cumsum(discharges)

# Plot results
plot(new_cases, type = "l")
lines(as.numeric(hospitalizations), col = "red")
lines(as.numeric(in_hospital), col = "blue")
```



## Summary

Summary text



::::::::::::::::::::::::::::::::::::: keypoints 

- Summarise the key points of the lesson using bullet points


::::::::::::::::::::::::::::::::::::::::::::::::
